Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-11-05 manu_draft.Rmd
reprocessed all models and summary data with model S

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(chgtrnd)
data(cmptrnd)
data(mods_chl18)
data(mods_chl24)
data(mods_chl27)
data(mods_chl32)
data(mods_chl34)

@@

This model is effectively equivalent to model S.  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.  With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.  With model SY, it is available more conveniently.  Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

Next, an average within-year cyclic pattern as a separate spline can be separted in the model formulation.  This can be expressed as:

@@

Models S, SY, and SYD can all potentially extract a similar signal from the raw data.  What differs is the allocation of penalties for curvature used to determine smoothness for each spline.  In model SYD, there are separate penalties for the two splines.  Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17]. For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations. In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.  The model will impute an average seasonal pattern where data are limited. However, a conceptual representation of what an average within-year pattern represents is challenging.  For example, suppose in every year includes a spring chlorophyll peak.  If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.  However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.  The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.

@@

Following similar rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.  Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.  The splines in `dec_time` and `doy` will likely not capture as much variation as above because of fewer knots.  The `ti` term represents an interaction by allowing the pattern in `dec_time` to vary by `doy` and vice-versa.  The interaction term in model SYDI provides an appearance that this model is fundamentally different form those provided by the other models.  However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.  Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually. By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.  However, this outcome is difficult to achieve for reasons explained below.

@@

p2 <- ggplot(toplo2, aes(x = mo, y = value, fill = yrcat)) + 
  geom_violin(draw_quantiles = 0.5)  + 
  scale_y_log10(ylab) +
  facet_wrap(~yrcat, ncol = 3) + 
  theme_minimal(base_family = 'serif') +
  scale_fill_manual(values = cls2) +
  theme(
    legend.title = element_blank(), 
    legend.position = 'none', 
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 40, hjust = 1), 
    strip.background = element_blank()
  ) +
  labs(
    y = ylab, 
    subtitle = '(b) Monthly concentrations by decade'
  )

>>>>

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(chgtrnd)
data(cmptrnd)
data(modslog_chl18)
data(modslog_chl24)
data(modslog_chl27)
data(modslog_chl32)
data(modslog_chl34)

@@

This model is effectively and mathematically equivalent to model S.  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.  With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.  With model SY, it is available more conveniently.  Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

Next, an average within-year cyclic pattern as a separate spline can be separated in the model formulation.  This can be expressed as:

@@

Models S, SY, and SYD can all potentially extract a similar signal from the raw data.  What differs between the models is the allocation of penalties for curvature used to determine smoothness for each spline.  In model SYD, there are separate penalties for the two splines, as compared to S and SY that include penalties only for the `cont_year` spline.  Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17]. For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations. In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.  The model will impute an average seasonal pattern where data are limited. However, a conceptual representation of what an average within-year pattern represents is challenging.  For example, suppose a spring chlorophyll peak is a notable feature in every year.  If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude thta appears as a "typical" year.  However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.  The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.

@@

Following similar rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.  Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.  The splines in `cont_year` and `doy` will likely not capture as much variation as above because of fewer knots.  The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.  The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.  However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.  Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually. By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.  However, this outcome is difficult to achieve for reasons explained below.

@@

p2 <- ggplot(toplo2, aes(x = mo, y = value, fill = yrcat)) + 
  geom_violin(draw_quantiles = 0.5)  + 
  scale_y_log10(ylab) +
  facet_wrap(~yrcat, ncol = 3) + 
  theme_minimal(base_family = 'serif') +
  scale_fill_manual(values = cls2) +
  theme(
    legend.title = element_blank(), 
    legend.position = 'none', 
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1), 
    strip.background = element_blank()
  ) +
  labs(
    y = ylab, 
    subtitle = '(b) Monthly concentrations by decade'
  )
